---
title: "Child Birthweight Data Representations"
author: "Kobbe Harwell"
date: ""
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r packages, echo = FALSE, message = FALSE}
library(ggplot2)
library(dplyr)
library(lattice)
```



Data Acquisition
```{r access}
site <- "http://www.stat.berkeley.edu/users/statlabs/data/babies.data"
BABIES <- read.table(file=url(site), header=TRUE)
summary(BABIES)
dim(BABIES)[2]
```


Removing subjects if any observations on the subject are "unknown". The variables `bwt`, `gestation`, `parity`, `age`, `height`, `weight`, and `smoke` use values of 999, 999, 9, 99, 99, 999, and 9, respectively, to denote ''unknown.''

```{r}
BABIES <- BABIES %>% mutate(bwt = ifelse(bwt == 999, NA, bwt))
BABIES <- BABIES %>% mutate(gestation = ifelse(gestation == 999, NA, gestation))
BABIES <- BABIES %>% mutate(parity = ifelse(parity == 9, NA, parity))
BABIES <- BABIES %>% mutate(age = ifelse(age == 99, NA, age))
BABIES <- BABIES %>% mutate(height = ifelse(height == 99, NA, height))
BABIES <- BABIES %>% mutate(weight = ifelse(weight == 999, NA, weight))
BABIES <- BABIES %>% mutate(smoke = ifelse(smoke == 9, NA, smoke))

```




Cleaning Data Set

```{r}
CLEAN <- na.omit(BABIES)
dim(CLEAN)
```



```{r}
colNAs<- colSums(is.na(BABIES))
colNAs
colNAs <- as.vector(colSums(is.na(BABIES)))

rowNAs <- table(rowSums(is.na(BABIES)))
rowNAs
rowNAs <- as.vector(table(rowSums(is.na(BABIES))))
```
>**Comment:**

                The missing values are as follows: 
                gestation: 13
                age: 2
                height: 22
                weight: 36
                smoke: 10
                1174 rows have no missing values, 43 rows have 1, 17 rows have 2, and 2 rows have 3 missing values.




Sorting the values in `CLEAN` by `bwt`, `gestation`, and `age`.  Storing the sorted values in a data frame named `byBGA`.

```{r}
byBGA <- CLEAN %>% arrange(bwt, gestation, age)
head(byBGA, 6)
tail(byBGA, 6)
```



Creating a new variable for bodyweight by grams.
```{r}

  CLEAN <- CLEAN %>% mutate(bwt_gr = bwt*28.35)
```



Removing 100 observation at random.
```{r}
set.seed(41)

remove <-sample(1:dim(CLEAN)[1],50)
print("Observation to remove:")
remove

CLEAN <- slice(CLEAN,-remove)
dim(CLEAN)[1]
head(CLEAN)
```



```{r save}
write.csv(CLEAN, file="CLEAN.csv", row.names=FALSE)
```






Finding the percentage of the women in `CLEAN` that are pregnant with their first child and do not smoke. Also those pregnant not for a first time and smoking.


```{r}
dim(filter(CLEAN, parity==0 & smoke ==0))[1]
dim(filter(CLEAN, parity==0 & smoke ==0))[1]/dim(CLEAN)[1]*100



dim(filter(CLEAN, parity==1 & smoke ==1))[1]
dim(filter(CLEAN, parity==1 & smoke ==1))[1]/dim(CLEAN)[1]*100
head(CLEAN)
```

> **Results:** 

                        # of women are pregnant with their first child and do not smoke.
                        ~45.3%
                        # of women not pregnant for first time and smoke:
                        ~10%



Creating a frequency histogram of the birth weights of babies whose mothers have never smoked and another histogram for the birth weights of babies whose mothers currently smoke.

```{r}

CLEAN <- CLEAN %>% mutate(smoke = factor(smoke, levels = 0:1, labels =c("Non-Smoker", "Smoker")))

   ggplot(data = CLEAN, aes(x =bwt_gr, y= ..density..)) +
   geom_histogram(fill = "lightblue") +
   geom_density(color = "orange") +
   facet_grid(smoke~.) +
   xlim(800, 5100) +
   xlab("Weight in Grams")+
   theme_bw()
```



Exploring Distribution

```{r}
# Using summarise() and group_by() from dplyr
  CLEAN %>% group_by(smoke) %>% summarize(mean(bwt_gr))
  CLEAN %>% group_by(smoke) %>% summarize(sd(bwt_gr))
  CLEAN %>% group_by(smoke) %>% summarize(var(bwt_gr))
  diff(range(CLEAN$bwt_gr))
```



Creating side-by-side boxplots to compare the birth weights (bwt_gr) of babies whose mother's never smoked and those who currently smoke.

```{r}
  ggplot(data=CLEAN, aes(x=smoke, y=bwt_gr)) +
  geom_boxplot()+
  ylab("Weight in Grams")

```




Finding the the median weight difference between babies who are firstborn and those who are not (`parity = 1`)?

```{r}
   CLEAN %>% group_by(parity) %>% summarize(median(bwt_gr))
```

> **Results:** The median weight difference between 
babies who are firstborn and those who are not is 99 grams.



Creating a graph of the densities for pre-pregnancy `weight` for mothers who have never smoked and for mothers who currently smoke. 
```{r}
    ggplot(data = CLEAN, aes(x =weight, color=smoke)) +
   geom_density(aes(fill = smoke), alpha=0.2) +
   xlab("Weight in Pounds")+
   theme_bw()
```



Computing the body mass index (`BMI`) for each mother in `CLEAN`. `BMI` is defined as $\text{kg}/\text{m}^2$.
```{r}
    CLEANP <- CLEAN %>% mutate(weightKg = weight*0.45359, heightM = height*0.0254, BMI = weightKg/(heightM^2))
    head(CLEANP)
```



Characterizing the distribution of BMI.

```{r}
   ggplot(data = CLEANP, aes(x =BMI)) +
   geom_density() +
    geom_vline(data=CLEANP, aes(xintercept = median(BMI)), linetype="dashed")+
   xlab("BMI (kg/m^2)")+
   theme_bw()
```

**Results:** 

The BMI distribution skews right and has a single local maximum.




Grouping pregnant mothers according to their BMI quartile.
```{r}
  cleanQuart <- CLEANP %>% mutate(BMIQuart = ntile(BMI, 4))
  head(cleanQuart %>% arrange(BMIQuart))
```


```{r Metrics}
    notSmoke <- cleanQuart %>% filter(smoke=="Non-Smoker") 
    notSmoke %>% group_by(BMIQuart) %>% summarize(mean(bwt_gr))
    notSmoke %>% group_by(BMIQuart) %>% summarize(sd(bwt_gr))
    
    notSmoke %>% group_by(BMIQuart) %>% summarize(median(bwt_gr))
    notSmoke %>% group_by(BMIQuart) %>% summarize(IQR(bwt_gr))
    
    
    didSmoke <- cleanQuart %>% filter(smoke=="Smoker")
    didSmoke %>% group_by(BMIQuart) %>% summarize(mean(bwt_gr))
    didSmoke %>% group_by(BMIQuart) %>% summarize(sd(bwt_gr))
    
    didSmoke %>% group_by(BMIQuart) %>% summarize(median(bwt_gr))
    didSmoke %>% group_by(BMIQuart) %>% summarize(IQR(bwt_gr))
```

```{r plotting}

    CLEANP <- CLEANP %>% mutate(BMIQuart = ntile(BMI, 4))

  ggplot(data = CLEANP, aes(x =bwt_gr)) +
   geom_density(color = "orange") +
   facet_grid(smoke~BMIQuart) +
    xlab("Weight in Grams")+
   theme_bw()
  
  ggplot(data = CLEANP, aes(x =bwt_gr)) +
   geom_histogram(fill = "lightblue") +
   facet_grid(smoke~BMIQuart) +
    xlab("Weight in Grams")+
   theme_bw()

```

> **Results: Birth weights in each group appear to be relatively symmetric.**




Creating boxplots of `bwt_gr` based on whether the mother smokes conditioned on `BMI` quartiles in order to determine Which one appears to be bigger factor, smoking or BMI?
```{r smokingVBMI}
  ggplot(data = CLEANP, aes(x=smoke, y= bwt_gr)) +
   geom_boxplot(color = "purple") +
   facet_grid(.~BMIQuart) +
    ylab("Weight in Grams")+
   theme_bw()

```

> **Results: Smoking appears to be a bigger factor on birth weight than BMI.**


```{r, echo=FALSE}
date()
sessionInfo()
R.Version()
```


